---
title: 'Group8: Zero-inflated Negative Binomial Regression'
author: "Amanda Lee Ketner, Yichao Chen, Xinrui Wu"
date: "12/10/2019"
output:
  html_document:
    toc: yes
  pdf_document: 
    latex_engine: xelatex
    toc: yes
subtitle: '*Analysis of the relationship between diet habit and alcohol use*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
library(foreign)
library(data.table)
library(knitr)
library(tidyverse)
library(pscl) # used for zero-inflated negative binomial regression
library(MASS) # used for negative binomial regression
library(ggplot2)
library(magrittr)
library(kableExtra)
library(Hmisc)
```

## Introduction
What is the relationship between eating habits and alcohol use?
Are certain types of eating habits associated with higher alcohol consumption?

Our group mainly focuses on the relationship between diet habit (specifically,
the self-reported healthfulness of a person's overall diet as well as
the average number of meals eaten outside the home weekly) and alcohol use 
(operationalized as the average number of alocholic drinks consumed weekly).
Other demographic variables (gender, family povery income ratio, and age) are also included as additional covariates.

Based on the distribution of the data, we chose to use zero-inflated negative binomial regression to fit the model. The choice of this regression is further illustrated in the method section.

## Data
### Data sets
We used three datasets in this analysis:

1.Alcohol Use (ALQ_D.xpt) 

2.Diet Behavior & Nutrition (DBQ_D.xpt) 

3.Demographic Variables & Sample Weights (DEMO_D.xpt) 

They are downloaded from NHANES (National Health and Nutrition Examination Survey) from year 2005 - 2006.

As we are making analysis on the alcohol use, so we will only focus on adults (age $\geqslant$ 21).

### Variables
```{r,echo=FALSE}
name=c('alq_drink','diet','meal_out','gender','age','pir')
ori_name = c("ALQ130", "DBQ700", "DBD091", "RIAGENDR", "RIDAGEYR", "INDFMPIR")
describe = c("Avg number of alcoholic drinks/day -past 12 mos (1-32:range of values)",
      "How healthy is the diet (1-5:Excellent-poor)", 
      "Number of times/wk eat meals not from a home (1-21:range of values)", 
      "Gender (1:Male 2:Female)",
      "Age at Screening Adjudicated - Recode (0-84:range of values)",
      "Family Poverty Income Ratio (0-5:a ratio of family income to poverty threshold)")
dataxpt=c('ALQ_D','DBQ_D','DBQ_D','DEMO_D','DEMO_D','DEMO_D')
data.table(Variales = name, Origin_name=ori_name,Description=describe,Dataset=dataxpt)%>%
  knitr::kable(align = "l") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), position = "left")



```

### Data cleaning process
We use three different methods to clean the data in the same process as follows. 

1. Select the variables needed and rename them.\newline
2. Drop observations with missing values. \newline
3. Since in the original data, people drink less than 1 time per day are recorded as 1, we replace 1 with 0 in the variable alq_drink. \newline
4. In the variable meal_out, approximate eating out "more than 21 times"/"less than 1 time" per week by "21 times"/"0 time". \newline
5. Select observations who are more than 21 years old.\newline
6. Merge the variables mentioned above together by the variable SEQN.\newline


## Methods
### Zero-inflated negative binomial regression
The zero-inflated negative binomial regression is used for count data that exhibit overdispersion and excess zeros. The data distribution combines the negative binomial distribution and the logit distribution.

Suppose for each observation, there are two possible cases. If case 1 occurs, the count is zero. If case 2 occurs, counts(including zeros) are generated according to the negative binomial model. We focus on our analysis here. Case 1 is that this person never drink because of some reasons such as alcohol allergy. Case 2 is that this person do have the habit of drink alcohol but the counts of drink may differ according to some related factors. Suppose that case 1 occurs with probability $\pi$ and case 2 occurs with probability $1-\pi$ . Therefore, the probability distribution of zero-inflated negative binomial regression random variable $y_{i}$ could be written:

$$
\operatorname{Pr}\left(y_{i}=j\right)=\left\{\begin{array}{ll}{\pi_{i}+\left(1-\pi_{i}\right) g\left(y_{i}=0\right)} & {\text { if } j=0} \\ {\left(1-\pi_{i}\right) g\left(y_{i}\right)} & {\text { if } j>0}\end{array}\right.
$$
where $\pi_{i}$ is the logistic link function defined below and $g\left(y_{i}\right)$ is the negative binomial distribution given by 
$$
g\left(y_{i}\right)=\operatorname{Pr}\left(Y=y_{i} | \mu_{i}, \alpha\right)=\frac{\Gamma\left(y_{i}+\alpha^{-1}\right)}{\Gamma\left(\alpha^{-1}\right) \Gamma\left(y_{i}+1\right)}\left(\frac{1}{1+\alpha \mu_{i}}\right)^{\alpha^{-1}}\left(\frac{\alpha \mu_{i}}{1+\alpha \mu_{i}}\right)^{y_{i}}
$$

Why we choose this zero-inflated negative binomial regression model to fit our data?

As observed, alq_drink is a count variable, this implies that the choice of possion regression or negative binomial regression may be more suitable. We then do some basic analysis of our response:alq_drink.It is overdispersed as its variance is not basiclly equal to its mean. In this case, negative binomial regression might be a better choice. Also, there are excessive zeros in alq_drink. As a result, zero-inflation should be taken into consideration as well. We will also make comparisions to figure out whether the consideration of zero-inflation could contribute to a better model. Details could be found in the core analysis below.

### Languages used
Amanda Lee Ketner: Stata \newline

Yichao Chen: R (including packages data.table and pscl) \newline

Xinrui Wu: SAS (for analysis) and R (with package dyplyr, for data cleaning) \newline

## Core Analysis {.tabset .tabset-fade}
### R
#### Data cleaning
The data cleaning process is completed using the package of data.table. 

```{r}
AL=read.xport('Data/ALQ_D.xpt')
write.csv(AL,file='Data/ALQ_D.csv')
DBQ=read.xport('Data/DBQ_D.xpt')
write.csv(DBQ,file='Data/DBQ_D.csv')
DEMO=read.xport('Data/DEMO_D.xpt')
write.csv(DEMO,file='Data/DEMO_D.csv')
ALQ_D= fread('Data/ALQ_D.csv')
DBQ_D=fread('Data/DBQ_D.csv')
DEMO_D=fread('Data/DEMO_D.csv')
# delete missing values of ALQ30 ane rename ALQ30 as alq_drink
# In the original data <1 drink  are recorded as 1, we replace 1 with 0 here
A1=ALQ_D[ALQ130!=''&ALQ130!=999,.(SEQN,alq_drink=ALQ130)][alq_drink==1,alq_drink:=0]
# delete missing values of DBD091 and DBQ700
# DBD091:5555(representing >21) consider as 21; 6666(representing <1) consider as 0
# rename DBD091 as meal_out; rename DBQ700 as diet
DBQ1=DBQ_D[DBD091!=''&DBD091!=7777&DBD091!=9999&DBQ700!=''&DBQ700!=7&DBQ700!=9][
  DBD091==6666,DBD091:=0][DBD091==5555,DBD091:=21][
    ,.(SEQN,meal_out=DBD091,diet=DBQ700)]
# delete missing values of RIAGENDR,RIDAGEYR,INDFMPIR
# only focus on adults(age>=21)
# rename RIAGENDR as gender;RIDAGEYR as age; INDFMPIR as pir
DEMO1=DEMO_D[RIAGENDR!=''&RIDAGEYR!=''&RIDAGEYR>=21&INDFMPIR!=''][
  RIAGENDR==2,RIAGENDR:=0][,.(SEQN,gender=RIAGENDR,age=RIDAGEYR,pir=INDFMPIR)]
#join these three datasets together according to SEQN
data=A1[DBQ1,on='SEQN',nomatch=0L][DEMO1,on='SEQN',nomatch=0L]
```

#### Basic data analysis
As discussed above, whether negative binomial regression is more suitable than poisson regression? Should the zero-inflation be considered? To figure this out, the basic analysis should be made on the response: alq_drink.

The mean and variance are calculated below:
```{r}
#calculate mean and variance
sprintf("Mean and Variance = %1.2f and %1.2f", 
        mean(data$alq_drink), var(data$alq_drink))
```

The plot of alq_drink are shown as below:
```{r}
# histogram with alq_drink
ggplot(data,aes(alq_drink))+geom_histogram()+stat_bin(bins=25)+xlab('alq_drink_counts')+ylab('frequency')+ggtitle('The frequency of alq_drink counts')+theme(plot.title=element_text(hjust=0.5))
```

From the result, alq_drink is overdispersed as its variance is not basiclly equal to its mean. So, the negative binomial regression is more suitable than poisson regression in our data. From the plot, we could see that a large part of observation has 0 alcohol drink. Taking excess zeros of alq_drink into consideration, zero-inflated regression is used to fit the data.

#### Zero-Inflated Negative Binomial Regression
The package of pscl is used for zero-inflated negative binomial regression. The variables of diet, gender, age and pir are used in the part of negative binomial model and the variable of meal_out is used in the logit part of the model.
```{r}
m1=zeroinfl(alq_drink~diet+gender+age+pir|meal_out,data=data,dist='negbin',EM=TRUE)
summary(m1)
m0=update(m1, . ~ 1)
pchisq(2 * (logLik(m1) - logLik(m0)), df = 5, lower.tail=FALSE)
```

From the output of chi-squared test, the p-value is quite small and we know that our overall model is statistically significant. 

From the result of m1, the variables of diet, gender, age and pir in the part of negative binomial are all significant predictors. The variable of meal_out in the part of the logit model predicting excessive zero is also statistically significant.

Holding other variable constant, the expeceted change in log(alq_drink) for one-unit increase in diet is 0.079589. From the codebook, the larger diet factor indicate poorer diet behavior. The model shows that poorer diet may related to more alcohol use.

When gender change from 0 to 1, the change in log(alq_drink) is 0.544, men tends to drink more than women.

The expeceted change in log(alq_drink) for one-unit increase in age is -0.016 holding other variable constant. This means when age increase, people tend to drink alcohol less.

The expeceted change in log(alq_drink) for one-unit increase in age is -0.097 holding other variable constant, which means family with better finacial situation might use alcohol less.

The log odds of being an excessive zero will decrease by 0.05516 for every one more meal eating outside. This means when the frequency of eating out of home is larger, the zero of alcohol use are less likely comes from the part of people who never use alcohol. In other words, more meals eating out of home may relate with more alcohol use.


#### Negative Binomial Regression
Has the consideration of zero-inflation improved our regression model? We could fit the data with negative binomial regression and make comparisions with the former model.

The package of MASS is used for building negative bionomial regression model.

```{r}
m2=glm.nb(alq_drink~diet+gender+age+pir,data=data)
summary(m2)
vuong(m1, m2)
```

The summary of m2 shows that the variables of diet, gender, age and pir are all significant. From the estimate of diet, gender, age and pir, we could know that poorer diet may related to more alcohol use. Men tends to drink more than women and family with better finacial situation might use alcohol less. The analysis process is quite similar to the analysis before in the zero-inflated negative binomial regression.

The result of Vuong test suggests that zero-inflated negative binomial regression have better performance and its improvement is significant.

### Stata
#### Cleaning the data
```{stata, echo=TRUE, comment = NA, eval = FALSE}

cd "/Data/"

*Import data
import sasxport5 "ALQ_D.xpt", clear
save "Alcohol.dta", replace
import sasxport5 "DBQ_D.xpt", clear
save "Diet.dta", replace
import sasxport5 "Demo_D.xpt", clear
save "Demographic.dta", replace

*Clean data and drop unwanted variables
use "Alcohol.dta", clear
keep seqn alq130
replace alq130=. if alq130==999
replace alq130=0 if alq130==1
rename alq130 alq_drink
save "Alcohol.dta", replace

use "Diet.dta", clear
keep seqn dbd091 dbq700
replace dbd091=0 if dbd091==6666
replace dbd091=21 if dbd091==5555
replace dbd091=. if dbd091==7777 | dbd091==9999
replace dbq700=. if dbq700==7 | dbq700==9
rename dbd091 meal_out
rename dbq700 diet
save "Diet.dta", replace

use "Demographic.dta", replace
keep seqn ridageyr riagendr indfmpir
replace riagendr=0 if riagendr==2
rename ridageyr age
rename riagendr gender
rename indfmpir pir
save "Demographic.dta", replace

*Merge data together and drop minors
use "Alcohol.dta", clear
merge 1:1 seqn using "Diet.dta", nogen
merge 1:1 seqn using "Demographic.dta", nogen
drop if Age<21
save "Final.dta", replace

```

#### Negative binomial model (without zero-inflation)

```{stata, echo=TRUE, comment = NA, eval = FALSE}

nbreg alq_drink diet i.gender age pir
outreg2 using NegBin.xls, stats(coef se tstat pval) noaster replace

```


```{r, echo=FALSE, comment = NA}

negbin_table = read.csv('Data/NegBin_clean.csv')

negbin_table %>% knitr::kable(caption = 'Non-Zero-Inflated Negative Binomial', digits=3)

```

#### Zero-Inflated Negative binomial model

```{stata, echo=TRUE, comment = NA, eval = FALSE}

zinb alq_drink diet i.gender age pir, inflate(meal_out) forcevuong
outreg2 using ZeroNegBin.xls, stats(coef se tstat pval) noaster replace

```

```{r, echo=FALSE, comment = NA}

negbin_table = read.csv('Data/ZeroNegBin_clean.csv')

negbin_table %>% knitr::kable(caption = 'Zero-Inflated Negative Binomial', digits=3)

```
Consistent with the results using the other methods, the Vuong test shows that 
the zero-inflated negative binomial model is preferred (p<0.000).

### SAS
#### Data cleaning
The cleaning process is done using R with the package dplyr.
```{r data1, results='hide'}
# read data :
alcohol = sasxport.get("./Data/ALQ_D.xpt")
diet = sasxport.get("./Data/DBQ_D.xpt")
demographic = sasxport.get("./Data/DEMO_D.xpt")

# select variables and merge : 
alcohol_need = transmute(alcohol, id = seqn, alq = alq130)
diet_need = transmute(diet, id = seqn, meal_out = dbd091, diet = dbq700)
demo_need = transmute(demographic, id = seqn, age = ridageyr, gender = riagendr, pir = indfmpir)
alq_diet = merge(alcohol_need, diet_need, by = "id")
alq_diet = merge(alq_diet, demo_need, by = "id")
```

```{r data2}
# replace "1" with "0" in alq
alq_diet[which(alq_diet$alq %in% c(777,999)),"alq"] = NA
alq_diet[which(alq_diet$alq == 1),"alq"] = 0
alq_diet[which(alq_diet$meal_out == 5555), "meal_out"] = 21
alq_diet[which(alq_diet$meal_out == 6666), "meal_out"] = 0
alq_diet[which(alq_diet$meal_out %in% c(7777,9999)),"meal_out"] = NA
alq_diet[which(alq_diet$gender == 2), "gender"] = 0
# remove all observations with missing values and select observations more the 21 years old
alq_diet = alq_diet%>%
  filter(age>=21) %>%
  filter(!is.na(alq) & !is.na(meal_out) & !is.na(pir) & !is.na(diet))

```

Write the new dataset to a csv file for later using in SAS.
```{r data3}
write.csv(alq_diet, file = "./SAS/alq_diet.csv")
```

#### Basic analysis
Read the new dataset into SAS. 
Check the distribution (mean and variance) of the variables.
```{sas read_data, eval = FALSE}
/* Read data */
proc import replace
	datafile='./alq_diet.csv' 
	out=alcohol_diet;
run;

proc print data=alcohol_diet;
run;

/* Basic checking */
proc means data = alcohol_diet mean var;
	var alq meal_out diet age gender pir;
run;

```

```{r check1, fig.align='center',out.width = 215, out.height = 122, echo=FALSE}
knitr::include_graphics("./SAS/dist_check.pdf")
```

Draw a histgram to see the distribution of the response variable ("alcohol use").
```{sas hist, eval = FALSE}
/* Frequency graph of alcohol */
ods graphics / width=4in height=3in border=off;
proc sgplot data=alcohol_diet;
	histogram alq /binwidth=1;
run;
ods graphics off;
```

```{r check2,  fig.align='center',out.width = 309, out.height = 224, echo=FALSE}
knitr::include_graphics("./SAS/hist_alq.pdf")
```

The response variable (alcohol use) is a count variable, so generally we can use Poisson regression. However, the mean and variance of the response shown above as "alq" are not the same, which does not fit the assumption of Poisson regression. So we choose to use negative binomial regression instead.

From the histgram we see that many people do not drink thus there are many 0's in the response. We can imagine that people's not drinking is influnced by another process compared to how much people drink. Thus we use zero-inflated negative binomial regression so that the 0's can be independently modeled.

#### Build Model1:negative binomial regression
All the following parts are done using SAS.\newline
Firstly, use negative binomial regression to fit model alq_drink ~ diet + gender + age + pir . The result is as below.\newline
```{sas nb, eval = FALSE}
/*  model analysis */
/* (1) negative binomial regression */
proc genmod data = alcohol_diet;
  model alq = diet gender age pir /dist=negbin;
run;

/* store the predict value of the model */
proc sort data = outnb;
  by id;
run;
```

```{r nbreg, echo=FALSE, fig.align='center',out.width = 609, out.height = 655,echo=FALSE}
knitr::include_graphics("./SAS/nb_results.pdf")
```

#### Build Model2: zero-inflated negative binomial regression
Then, use zero-inflated negative binomial regression to fit model alq_drink ~ diet + gender + age + pir where alq_drink is not 0 and fit model alq_drink ~ meal_out where alq_drink is 0. The result is as below.\newline
```{sas zinb, eval = FALSE}
/* (2) zero-inflated negative binomial regression */
proc genmod data = alcohol_diet order = data;
  model alq = diet gender age pir /dist=zinb;
  zeromodel meal_out;
run;

/* store the predict value of the model */
proc sort data = outzinb;
  by id;
run;
/* merge the predict values of the two model */
data predict;
  merge outnb outzinb;
  by id;
run;
```

```{r zinbreg, echo=FALSE, fig.align='center',out.width = 609, out.height = 810,echo=FALSE}
knitr::include_graphics("./SAS/zinb_results.pdf")
```

We can see from the results that in both models, all the variables are significant. By the AIC value, we can roughly see that the zero-inflated negative binomial model is better. \new

#### Use vuong test to compare the two models
More specificly, we use vuong test to compare result of the two models.
```{sas vuong, eval = FALSE}
%include "./vuong.sas";
%vuong(data = predict, response = alq,
	   model1 = nb,   p1 = prednb, dist1 = nb, scale1 = 0.6381,
	   model2 = zinb, p2 = predzinb, dist2 = zinb, scale2 = 0.2236, pzero2 = p0,
       nparm1 = 4, nparm2 = 5)
```

```{r vuong result, echo=FALSE, fig.align='center',out.width = 607, out.height = 740,echo=FALSE}
knitr::include_graphics("./SAS/vuong_results.pdf")
```

We can see from the test that the zero-inflated negative binomial regression model is preffered.\newline

## Additional analysis
### Predict alq_drink and Show results in plot
We could compute the predicted number of alcohol use for diffrent combinations of our predictors. The plot of predicted alq_drink counts and variables may help us see the potential relationship more directly.
```{r}
newdata1 <- expand.grid(1:5, factor(0:1),21:85,0:5,1:21)
colnames(newdata1)<-c('diet','gender','age','pir','meal_out')
newdata1$alqpre<-predict(m1,newdata1)
ggplot(newdata1, aes(x = diet, y = alqpre, colour = factor(pir))) +
  geom_point() +
  facet_wrap(~gender) +
  labs(x = "diet", y = "Predicted alcohol drink counts")
```

From the plot, we could see directly that male use alcohol more than women. Family with smaller  poverty income ratio, which means family in poorer financial situation have larger predicted  alcohol drinking counts. And people with unheathier diet tend to have larger drinking counts as well. This corresonds with our illustration in the part of core analysis that the use of alchhol may have some correlation with gender, family PIR and diet habit.

## Results

Overall, we found a statistically significant and positive relationship between the self-reported
healthfulness of a person's diet and their alcoholic consumption. In other words, people with less healthy diets tend to drink more alcohol. \newline

The use of three different languages: Stata, R and SAS reflect basically the same results and some detailed explanations of our zero-inflated negative binomial regression are as follows:

1. The expeceted change in log(alq_drink) for one-unit increase in diet is about 0.08 holding other variables constant. Since, from the codebook, the larger diet factor indicate poorer diet behavior, we have evidence to say that people who drink and have healthier diet tend to drink more. \newline

2. With other variables kept the same, log(alq_drink) is approximately 0.5 larger when gender = 1(male) than gender = 0(female), indicating that among drinking people, men tend to drink more than women do. \newline

3. The expected change in log(alq_drink) for one-unit increase in age is -0.02 holding other variables constant. This means among drinking people, older people tend to drink less alcohol. \newline

4. The expected change in log(alq_drink) for one-unit increase in family poverty income ratio is -0.10 holding other variables constant, which means that means among drinking families, wealtheir families consume less alcohol on average. \newline

5. The log odds of being an excessive zero will decrease by 0.06 for every one more meal eating outside. This means when the frequency of eating out of home is larger, the zero are less likely comes from the part of people who never use alcohol. In other words, more meals eating out of home implies more alcohol use. \newline

Besides, using Stata, R and SAS, we all find zero-inflated negative binomial regression works better for this data than negative binomial regression does. \newline

## References

https://ncss-wpengine.netdna-ssl.com/wp-content/themes/ncss/pdf/Procedures/NCSS/Zero-Inflated_Negative_Binomial_Regression.pdf

https://www.stata.com/manuals13/rzinb.pdf

https://stats.idre.ucla.edu/r/dae/zinb/

https://stats.idre.ucla.edu/sas/dae/zero-inflated-negative-binomial-regression/

