---
title: "Group8 Alcohol Use and Eating Habit"
author: "Xinrui Wu"
date: "12/5/2019"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library, include=FALSE, results='hide'}
library(Hmisc)
library(tidyverse)
```

## Introduction
Our group mainly focus on the relationship between eating habit and alcohol use of people while some demographic variables are also included. Zero inflation negative binomial regression model is used in our analysis process due to the structure of the response variable.

## Data to Use
Three datasets from NHANES 2005-2006  are used:\newline
1.Alcohol Use (ALQ_D.xpt)\newline
2.Diet Behavior & Nutrition (DBQ_D.xpt)\newline
3.Demographic Variables & Sample Weights (DEMO_D.xpt)\newline

From these three datasets, we choose the following variable:
ALQ130 - Avg # alcoholic drinks/day -past 12 mos (1-32:range of values) - "alcohol"\newline
DBQ700 - How healthy is the diet (1-5:Excellent-poor) - "diet"\newline
DBD091 - # of times/wk eat meals not from a home (1-21:range of values) - "meal_out" \newline
RIAGENDR - Gender (1:Male 2:Female) - "gender" \newline
RIDAGEYR - Age at Screening Adjudicated - Recode (0-84:range of values) - "age" \newline
INDFMPIR - Family PIR (0-5:a ratio of family income to poverty threshold) - "pir" \newline

Here we only focus on adults and select samples with age>=21.

## Programming Method
R (using dplyr for basic data cleaning) and SAS

## 1. Data Cleaning
The cleaning process is done using R with the package dplyr.
Read the data the join variables to use in a new dataset.
```{r data1}
# read data :
alcohol = sasxport.get("./ALQ_D.xpt")
diet = sasxport.get("./DBQ_D.xpt")
demographic = sasxport.get("./DEMO_D.xpt")

# select variables and merge : 
alcohol_need = transmute(alcohol, id = seqn, alq = alq130)
diet_need = transmute(diet, id = seqn, meal_out = dbd091, diet = dbq700)
demo_need = transmute(demographic, id = seqn, age = ridageyr, gender = riagendr, pir = indfmpir)
alq_diet = merge(alcohol_need, diet_need, by = "id")
alq_diet = merge(alq_diet, demo_need, by = "id")
```

Since in the raw dataset, people who don't drink report "1" in the variable "ALQ130", transfer 1's in the related variable to 0's.
Delete all missing values.

```{r data2}
alq_diet[which(alq_diet$alq %in% c(777,999)),"alq"] = NA
alq_diet[which(alq_diet$alq == 1),"alq"] = 0
alq_diet[which(alq_diet$meal_out == 5555), "meal_out"] = 21
alq_diet[which(alq_diet$meal_out == 6666), "meal_out"] = 0
alq_diet[which(alq_diet$meal_out %in% c(7777,9999)),"meal_out"] = NA
alq_diet[which(alq_diet$gender == 2), "gender"] = 0
alq_diet = alq_diet%>%
  filter(age>=21) %>%
  filter(!is.na(alq) & !is.na(meal_out) & !is.na(pir) & !is.na(diet))

```

Write the new dataset to a csv file for later using in SAS.
```{r data3}
write.csv(alq_diet, file = "alq_diet.csv")
```

## 2.Basic analysis
Read the new dataset into SAS. 
Check the distribution (mean and variance) of the variables.
```{sas read_data, eval = FALSE}
/* Read data */
proc import replace
	datafile='./alq_diet.csv' 
	out=alcohol_diet;
run;

proc print data=alcohol_diet;
run;

/* Basic checking */
proc means data = alcohol_diet mean var;
	var alq meal_out diet age gender pir;
run;

```

```{r check1, fig.align='center',out.width = 215*3, out.height = 122*3, echo=FALSE}
knitr::include_graphics("./dist_check.pdf")
```

Draw a histgram to see the distribution of the response variable ("alcohol use").
```{sas hist, eval = FALSE}
/* Frequency graph of alcohol */
ods graphics / width=4in height=3in border=off;
proc sgplot data=alcohol_diet;
	histogram alq /binwidth=1;
run;
ods graphics off;
```

```{r check2,  fig.align='center',out.width = 309*2, out.height = 224*2, echo=FALSE}
knitr::include_graphics("./hist_alq.pdf")
```

The response variable (alcohol use) is a count variable, so generally we can use Poisson regression. However, the mean and variance of the response shown above as "alq" are not the same, which does not fit the assumption of Poisson regression. So we choose to use negative binomial regression instead.

From the histgram we see that many people do not drink thus there are many 0's in the response. We can imagine that people's not drinking is influnced by another process compared to how much people drink. Thus we use zero-inflated negative binomial regression so that the 0's can be independently modeled.

## 3.Build Model
The following process is done using SAS.\newline
#### negative binomial regression
Firstly, use negative binomial regression to fit model alcohol ~ diet + gender + age + pir . The result is as below.\newline
```{sas nb, eval = FALSE}
/*  model analysis */
/* (1) negative binomial regression */
proc genmod data = alcohol_diet;
  model alq = diet gender age pir /dist=negbin;
run;

/* store the predict value of the model */
proc sort data = outnb;
  by id;
run;
```

```{r nbreg, echo=FALSE, fig.align='center',out.width = 609, out.height = 655,echo=FALSE}
knitr::include_graphics("./nb_results.pdf")
```

#### zero-inflated negative binomial regression
Then, use zero-inflated negative binomial regression to fit model alcohol ~ diet + gender + age + pir where alcohol is not 0 and fit model alcohol ~ meal_out where alcohol is 0. The result is as below.\newline
```{sas zinb, eval = FALSE}
/* (2) zero-inflated negative binomial regression */
proc genmod data = alcohol_diet order = data;
  model alq = diet gender age pir /dist=zinb;
  zeromodel meal_out;
run;

/* store the predict value of the model */
proc sort data = outzinb;
  by id;
run;
/* merge the predict values of the two model */
data predict;
  merge outnb outzinb;
  by id;
run;
```

```{r zinbreg, echo=FALSE, fig.align='center',out.width = 609, out.height = 810,echo=FALSE}
knitr::include_graphics("./zinb_results.pdf")
```

#### Use vuong test to compare the two models
```{sas vuong, eval = FALSE}
%include "./vuong.sas";
%vuong(data = predict, response = alq,
	   model1 = nb,   p1 = prednb, dist1 = nb, scale1 = 0.6381,
	   model2 = zinb, p2 = predzinb, dist2 = zinb, scale2 = 0.2236, pzero2 = p0,
       nparm1 = 4, nparm2 = 5)
```

```{r vuong result, echo=FALSE, fig.align='center',out.width = 607, out.height = 740,echo=FALSE}
knitr::include_graphics("./vuong_results.pdf")
```

We can see from the test that the zero-inflated negative binomial regression model is preffered.\newline



