---
title: "Group8_R"
author: "Yichao Chen"
date: "12/3/2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
library(foreign)
library(data.table)
library(knitr)
library(pscl) # used for zero-inflated negative binomial regression
library(MASS) # used for negative binomial regression
library(ggplot2)
```

# R
This file is the part of analysis process using R in Group8 Project.

## Data Cleaning
Firstly, the datasets are cleaned. All the missing values are deleted and three datasets are joined. Here, the package of data.table is used. Details could be seen in group8.Rmd and comments. In this process, we rename the variables for convenience: ALQ130:alq_drink, DBD091: meal_out, DBQ700:diet, RIAGENDR :gender, RIDAGEYR:age, INDFMPIR:pir.

```{r}
AL=read.xport('Data/ALQ_D.xpt')
write.csv(AL,file='Data/ALQ_D.csv')
DBQ=read.xport('Data/DBQ_D.xpt')
write.csv(DBQ,file='Data/DBQ_D.csv')
DEMO=read.xport('Data/DEMO_D.xpt')
write.csv(DEMO,file='Data/DEMO_D.csv')
ALQ_D= fread('Data/ALQ_D.csv')
DBQ_D=fread('Data/DBQ_D.csv')
DEMO_D=fread('Data/DEMO_D.csv')
# delete missing values of ALQ30 ane rename ALQ30 as alq_drink
# In the original data <1 drink  are recorded as 1, we replace 1 with 0 here
A1=ALQ_D[ALQ130!=''&ALQ130!=999,.(SEQN,alq_drink=ALQ130)][alq_drink==1,alq_drink:=0]
# delete missing values of DBD091 and DBQ700
# DBD091:5555(representing >21) consider as 21; 6666(representing <1) consider as 0
# rename DBD091 as meal_out; rename DBQ700 as diet
DBQ1=DBQ_D[DBD091!=''&DBD091!=7777&DBD091!=9999&DBQ700!=''&DBQ700!=7&DBQ700!=9][
  DBD091==6666,DBD091:=0][DBD091==5555,DBD091:=21][
    ,.(SEQN,meal_out=DBD091,diet=DBQ700)]
# delete missing values of RIAGENDR,RIDAGEYR,INDFMPIR
# only focus on adults(age>=21)
# rename RIAGENDR as gender;RIDAGEYR as age; INDFMPIR as pir
DEMO1=DEMO_D[RIAGENDR!=''&RIDAGEYR!=''&RIDAGEYR>=21&INDFMPIR!=''][
  RIAGENDR==2,RIAGENDR:=0][,.(SEQN,gender=RIAGENDR,age=RIDAGEYR,pir=INDFMPIR)]
#join these three datasets together according to SEQN
data=A1[DBQ1,on='SEQN',nomatch=0L][DEMO1,on='SEQN',nomatch=0L]
```


## Basic data analysis
As discussed above, whether negative binomial regression is more suitable than poisson regression? Should the zero-inflation be considered? To figure this out, the basic analysis should be made on the response: alq_drink.

The mean and variance are calculated below:

```{r}
#calculate mean and variance
sprintf("Mean and Variance = %1.2f and %1.2f", 
        mean(data$alq_drink), var(data$alq_drink))
```

The plot of alq_drink are shown as below:

```{r}
# histogram with alq_drink
ggplot(data,aes(alq_drink))+geom_histogram()+stat_bin(bins=25)+xlab('alq_drink_counts')+ylab('frequency')+ggtitle('The frequency of alq_drink counts')+theme(plot.title=element_text(hjust=0.5))
```

From the result, alq_drink is overdispersed as its variance is not basiclly equal to its mean. So, the negative binomial regression is more suitable than poisson regression in our data. From the plot, we could see that a large part of observation has 0 alcohol drink. Taking excess zeros of alq_drink into consideration, zero-inflated regression is used to fit the data.

## Zero-Inflated Negative Binomial Regression
The package of pscl is used for zero-inflated negative binomial regression. The variables of diet, gender, age and pir are used in the part of negative binomial model and the variable of meal_out is used in the logit part of the model.

```{r}
m1=zeroinfl(alq_drink~diet+gender+age+pir|meal_out,data=data,dist='negbin',EM=TRUE)
summary(m1)
m0=update(m1, . ~ 1)
pchisq(2 * (logLik(m1) - logLik(m0)), df = 5, lower.tail=FALSE)
```
From the output of chi-squared test, the p-value is quite small and we know that our overall model is statistically significant. 

From the result of m1, the variables of diet, gender, age and pir in the part of negative binomial are all significant predictors. The variable of meal_out in the part of the logit model predicting excessive zero is also statistically significant.

Holding other variable constant, the expeceted change in log(alq_drink) for one-unit increase in diet is 0.079589. From the codebook, the larger diet factor indicate poorer diet behavior. The model shows that poorer diet may related to more alcohol use.

When gender change from 0 to 1, the change in log(alq_drink) is 0.544, men tends to drink more than women.

The expeceted change in log(alq_drink) for one-unit increase in age is -0.016 holding other variable constant. This means when age increase, people tend to drink alcohol less.

The expeceted change in log(alq_drink) for one-unit increase in age is -0.097 holding other variable constant, which means family with better finacial situation might use alcohol less.

The log odds of being an excessive zero will decrease by 0.05516 for every one more meal eating outside. This means when the frequency of eating out of home is larger, the zero of alcohol use are less likely comes from the part of people who never use alcohol. In other words, more meals eating out of home may relate with more alcohol use.


## Negative Binomial Regression
Has the consideration of zero-inflation improved our regression model? We could fit the data with negative binomial regression and make comparisions with the former model.

The package of MASS is used for building negative bionomial regression model.

```{r}
m2=glm.nb(alq_drink~diet+gender+age+pir,data=data)
summary(m2)
vuong(m1, m2)
```

The summary of m2 shows that the variables of diet, gender, age and pir are all significant. From the estimate of diet, gender, age and pir, we could know that poorer diet may related to more alcohol use. Men tends to drink more than women and family with better finacial situation might use alcohol less. The analysis process is quite similar to the analysis before in the zero-inflated negative binomial regression.

The result of Vuong test suggests that zero-inflated negative binomial regression have better performance and its improvement is significant.

## Additional analysis
### Predict alq_drink and Show results in plot

We could compute the predicted number of alcohol use for diffrent combinations of our predictors. The plot of predicted alq_drink counts and variables may help us see the potential relationship more directly.

```{r}
newdata1 <- expand.grid(1:5, factor(0:1),21:85,0:5,1:21)
colnames(newdata1)<-c('diet','gender','age','pir','meal_out')
newdata1$alqpre<-predict(m1,newdata1)
ggplot(newdata1, aes(x = diet, y = alqpre, colour = factor(pir))) +
  geom_point() +
  facet_wrap(~gender) +
  labs(x = "diet", y = "Predicted alcohol drink counts")
```

From the plot, we could see directly that male use alcohol more than women. Family with smaller  poverty income ratio, which means family in poorer financial situation have larger predicted  alcohol drinking counts. And people with unheathier diet tend to have larger drinking counts as well. This corresonds with our illustration in the part of core analysis that the use of alchhol may have some correlation with gender, family PIR and diet habit.
